Species persistence under climate and fishing

### 2014-01-23 Writing/MS.tex
Finished intro edits, this closes #1

There are many stressors that can disturb an ecosystem. Ecologists have quantified the effects of a number of stressors individually \citep{Wilcoveetal1998, Crainetal2008, DarlingCote2008}. However, disturbances rarely occur in isolation.  It is therefore important to understand how a system is affected by multiple disturbances together \citep{DoakMorris2010, Fordhametal2013, Foltetal1999}. A perturbation that has little effect when affecting a system individually may amplify the disturbance caused by a coincident perturbation \citep{Crainetal2008, DarlingCote2008}. Synergistic interactions between stressors are especially worrying if an animal population can persist in the face of a single stressor but not in the presence of multiple stressors (e.g.  \citet{Pelletieretal2006}). However, it remains hard to predict when disturbances are likely to interact in ways not predicted by the effects of individual perturbations.

Climate change is a prime example of a stressor occurring in a background of multiple other, possibly interacting, disturbances. These include pollution, habitat fragmentation, invasive species, and over-exploitation \citep{Wilcoveetal1998, Salaetal2000, MEA2005}. It is expected that  species will move their home ranges in response to climate change. Many marine fish are already moving \citep{Perryetal2005, HiddinkHoftstede2008, Rijnsdorpetal2009, Dulvyetal2008, Simpsonetal2011} and are projected to continue to move in the future \citep{Kelletal2005, Mackenzieetal2007}. These species are faced not only with temperature-driven range range shifts but also with anthropogenic stressors including ocean acidification and commercial fishing \citep{Pinskyetal2013, Barryetal1995, Nyeetal2009}. Climate change and fishing have been identified as the two largest human impacts on the ocean \citep{Halpernetal2008} so there is a need to understand whether and how these disturbances will interact in the future.

Many methods have been proposed to help mitigate the effects of harvesting on fish populations. Marine protected areas (MPAs), which provide stepping stones to help species keep up with a changing environment, have been suggested as a key form of climate insurance \citep{Thomasetal2012, Hannahetal2007}. MPAs are frequently recommended for both conservation of biodiversity and improved fisheries yield \citep{Gainesetal2010}. Threshold harvesting levels below which no fish may be harvested have also been proposed as a means of both optimizing yield and conserving resources \citep{Landeetal1995, Landeetal1997}.

A wide range of models have been used to study the effect of climate
change and harvesting on fish populations.  A common approach to predicting future population distributions under climate change has been to use bioclimatic-envelope models (also known as species distribution models -- SDMs). These statistical models have been
used to find correlations between the presence or absence of fish
species and biophysical properties of the environment such as mean or maximum
temperatures, rainfall, or salinity, to explain and predict how species
ranges will differ under climate change \citep{Elithetal2006, GuisanThuiller2005, GuisanZimmerman2000}. Similarly,
projection models of future climate conditions have been used to predict
where conditions will be favorable for fish populations in the future
\citep{Cheungetal2008, Cheungetal2009}.  Despite these models' widespread
adoption, they have  been criticized for being oversimplified
as they lack species interactions, dispersal and reproductive processes
\citep{KearneyPorter2009, Zarnetskeetal2012, Robinsonetal2011}.

Some recent work on range shifts has addressed these gaps by explicitly including dispersal and reproduction \citep{Berestyckietal2009, ZhouKot2011}, which have been highlighted as important factors in fish populations' vulnerability to climate change \citep{Fordhametal2013, Hastingsetal2005}.  These models only address one disturbance, climate-driven range shifts, and do not consider other anthropogenic stressors.  The models of Zhou and Kot
\citep{ZhouKot2011} and Berestycki et al. \citep{Berestyckietal2009} are of
intermediate complexity (sensu
\citep{Gaylordetal2005}).  Work that consider both climate-driven
range shifts and fishing is often case-specific and detailed, including
multiple drivers and disturbances \citep{Cheungetal2010, Lindegrenetal2010, Brownetal2010, Merinoetal2010, Merinoetal2010b, Plaganyietal2011, Ainsworthetal2011, Zhangetal2011,
Barangeetal2011, Howardetal2013}. These cumulative
impacts are important for management and conservation planning
\citep{Allisonetal2009}.  However, these models are
so complex that understanding the relative importance of particular
drivers, disturbances, and interactions is difficult (but see
\citep{Nyeetal2013} for an approach using ecosystem-level models to discern
relative importance of disturbances).

We analyzed an integrodifference model to study the effect of temperature-driven range shifts and
harvesting, two important interacting disturbances, on the persistence of a marine population. In particular, we extended the model of
Zhou and Kot \citep{ZhouKot2011}, which describes a fish population
reproducing and dispersing while their range shifts, by
additionally including harvesting pressure and introducing two different
conservation measures. We found that the disturbances interact
synergistically. We also found that MPAs can help a species persist with
higher harvesting pressure, but they do not change the maximum speed of
habitat shift with which a species can keep up.

>>>>

There are many stressors that can disturb an ecosystem. Ecologists have quantified the effects of a number of stressors individually \citep{Wilcoveetal1998, Crainetal2008, DarlingCote2008}. However, disturbances rarely occur in isolation and it is therefore important to understand how a system is affected by multiple disturbances together \citep{DoakMorris2010, Fordhametal2013, Foltetal1999}. A perturbation that has little effect when it occurs individually may still amplify the disturbance caused by a coincident perturbation \citep{Crainetal2008, DarlingCote2008}. Synergistic interactions between stressors can be especially problematic if an animal population can persist in the face of a single stressor but the combination of multiple stressors suffices to drive the population extinct (i.e. \citet{Pelletieretal2006}). However it remains difficult to predict when disturbances are likely to interact in ways not anticipated when perturbations are measured individually.

Climate change is a prime example of multiple, possibly interacting, disturbances. Species that are likely to undergo (or already undergoing) shifts in range and phenology are subject to many other disturbances including pollution, habitat fragmentation, over-exploitation, and invasive species \citep{Wilcoveetal1998, Salaetal2000, MEA2005}. A common approach to predicting future population distrubtions under climate change has been to use bioclimatic-envelope models (also known as species distribution models -- SDMs). These statistical models typically correlate presence-absence data with biophysical characteristics such as mean or maximum temperatures, rainfall, or salinity, to explain and predict how species ranges' will differ under climate change \citep{Elithetal2006, GuisanThuiller2005, GuisanZimmerman2000}. Despite these models' widespread adoption, SDMs have frequently been criticized as oversimplified as they lack species interactions, dispersal and reproductive processes \citep{KearneyPorter2009, Zarnetskeetal2012, Robinsonetal2011}. Recent work on range shifts has addressed some of these gaps by explicitly including dispersal and reproduction \citep{Berestyckietal2009, ZhouKot2011}. However these models only address one disturbance: that of climate-driven range shifts. Not only does climate change disturb ecological systems via range shifts (for example), but these range shifts may interact with anthropogenetic stressors already present. Marine fish, for example, are already moving \citep{Perryetal2005, HiddinkHoftstede2008, Rijnsdorpetal2009, Dulvyetal2008, Simpsonetal2011} and projected to move more  \citep{Kelletal2005, Mackenzieetal2007}. These organisms not only are faced with ocean acidification and temperature driven range shifts \citep{Pinskyetal2013, Barryetal1995, Nyeetal2009}, but are often harvested commercially.

Climate change and fishing have been identified as the two largest human impacts on the ocean \citep{Halpernetal2008} so there is a need to understand whether and how these disturbances will interact in the future. Predictions for how climate change will drive shifts in marine systems have focused on how physiological limitations \citep{PortnerKnust2007} and global fish distributions \citep{Cheungetal2008, Cheungetal2009, Hareetal2010, Lenoiretal2011} may change. Work considering the joint impacts of climate and fishing often consider climate fluctuations (large anomalies around the mean) rather than directional changes in climate \citep{WaltersParma1996, KingMcFarlane2006}. Work that does consider the effects of climate-driven range shifts on fishing typically are case-specific and detailed, integrating multiple drivers and disturbances \citep{Cheungetal2010, Lindegrenetal2010, Brownetal2010, Merinoetal2010, Merinoetal2010b, Plaganyietal2011, Ainsworthetal2011, Zhangetal2011, Barangeetal2011, Howardetal2013}. These predicted impacts are important for management and conservation planning \citep{Allisonetal2009}, however these models are so complex that understanding the relative importance of particular drivers, disturbances, and interactions is difficult (but see \citet{Nyeetal2013} for an approach using ecosystem-level models to discern relative importance of disturbances). Here we develop models of intermediate complexity \citep[sensu][]{Gaylordetal2005} that can help clarify the importance of two interacting disturbances: temperature-driven range shifts and harvest. Empirical case studies have identified interactions between overfishing and temperature-driven range shifts that suggest synergy in the magnitude of the disturbance \citep{Lingetal2009} and have demonstrated synergistic effects in micosm experiments between warming temperatures, harvesting and connectivity \citep{Moraetal2007}. This empirical work underscores the importance of understanding how range shifts and harvesting interact and, to our knowledge, there has been limited work examining interactions between harvesting and range shifts.

Here we construct an analytical model of a fish population subject to both harvesting and climate change-induced range-shift. We explicitly include two mechanistic processes central to species' responses to climate and fishing: reproduction and dispersal. Previous work has highlighted the importance of these two processes and their vulnerability to climate change \citep{Fordhametal2013, Hastingsetal2005}. We focus on determining how climate-driven range shifts and fishing interact to affect population persistence. We find that these disturbances interact additively, not multiplicatively which is often the default assumption when considering mortality.

We also examine the effect of marine protected areas (MPAs) on species persistence. Protected areas have been suggested as a key form of climate insurance and stepping stones to help species keep up with a changing environment \citep{Thomasetal2012, Hannahetal2007}. MPAs are frequently recommended for conservation of biodiversity and improved fisheries yield \citep{Gainesetal2010}, and we evaluate whether MPAs established for those purposes could improve species persistence when habitat is shifting rapidly. We find that MPAs can help a species persist with higher harvesting pressure, but does not change the maximum climate velocity with which a species can keep up.
